---
title: "Issues on DAS indicator"
author: "Els Lommelen"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(forrescalc)
library(dplyr)
library(tidyr)
library(knitr)
library(kableExtra)
library(readr)
library(ggplot2)
path_to_fieldmap <- "C:/3BR/2_VisualisatieDataBR/3MDB_BOSRES_selectieEls/FieldMapData_MDB_BOSRES_selectieEls.accdb"
```

# Selection of forest reserves and plots

Following the method of Bosch & Partner (2014), selection of forest reserves based on the 3 criteria should be done in 2 steps:

- select the forest reserves that meet these criteria
- select the plots in the reserves that meet these criteria

But is it necessary to do this in 2 steps? Is there any difference in selected between doing 2 steps and doing only the last step?

## The 2-step method:

```{r results="asis"}
data_dendro <- load_data_dendrometry(path_to_fieldmap) %>%
  filter(alive_dead == 11)
selection_2steps <- data_dendro %>%
  select_for_DAS_indicator(
    grouping_vars = c("forest_reserve", "year", "period")
  ) %>%
  inner_join(data_dendro, by = c("forest_reserve", "year", "period")) %>%
  select_for_DAS_indicator(grouping_vars = c("plot_id", "year", "period")) %>%
  inner_join(
    data_dendro %>%
      select(plot_id, year, period, forest_reserve) %>%
      distinct(),
    by = c("plot_id", "year", "period")
  )

selection_2steps %>%
  group_by(forest_reserve, period) %>%
  summarise(n_plots = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = period,
              names_prefix = "period",
              values_from = n_plots) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

```{r results="asis"}
selection_2steps %>%
  pivot_wider(names_from = period,
              names_prefix = "period",
              values_from = year) %>%
  group_by(forest_reserve) %>%
  summarise(
    n_plots_period1 = sum(!is.na(period1) & is.na(period2)),
    n_plots_period2 = sum(is.na(period1) & !is.na(period2)),
    n_plots_period12 = sum(!is.na(period1) & !is.na(period2))
  ) %>%
  ungroup() %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```


## Only the last step:

```{r results="asis"}
selection_1step <- data_dendro %>%
  select_for_DAS_indicator(grouping_vars = c("plot_id", "year", "period")) %>%
  inner_join(
    data_dendro %>%
      select(plot_id, year, period, forest_reserve) %>%
      distinct(),
    by = c("plot_id", "year", "period")
  )

selection_1step %>%
  group_by(forest_reserve, period) %>%
  summarise(n_plots = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = period,
              names_prefix = "period",
              values_from = n_plots) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

```{r results="asis"}
selection_1step %>%
  pivot_wider(names_from = period,
              names_prefix = "period",
              values_from = year) %>%
  group_by(forest_reserve) %>%
  summarise(
    n_plots_period1 = sum(!is.na(period1) & is.na(period2)),
    n_plots_period2 = sum(is.na(period1) & !is.na(period2)),
    n_plots_period12 = sum(!is.na(period1) & !is.na(period2))
  ) %>%
  ungroup() %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

## Calculation of criteria on reserve level (for omitted reserves)

```{r results="asis", message=FALSE}
data_dendro %>%
  group_by(forest_reserve, year, period) %>%
  summarise(
    dbh_mm_average = mean(.data$dbh_mm)
  ) %>%
  ungroup() %>%
  left_join(data_dendro, by = c("forest_reserve", "year", "period")) %>%
  left_join(
    read_csv2(system.file("./extdata/DAS_tree_groups.csv",
                          package = "forrescalc")),
    by = "species"
  ) %>%
  group_by(forest_reserve, year, period, dbh_mm_average, group) %>%
  summarise(
    basal_area_m2_ha = sum(.data$basal_area_alive_m2_ha)
  ) %>%
  ungroup() %>%
  group_by(forest_reserve, year, period, dbh_mm_average) %>%
  mutate(
    basal_area_proportion =
      .data$basal_area_m2_ha / sum(.data$basal_area_m2_ha)
  ) %>%
  ungroup() %>%
  kable()

```

## Conclusie

Heirnisse valt weg doordat de gemiddelde diameter van de bomen te laag is.

Ander aandachtspunt voor dit reservaat (moest het wel meegenomen kunnen worden): de plots zijn niet allemaal in dezelfde winter opgemeten. Zijn er nog reservaten waarbij dit het geval is? Hoe gaan we de berekening in dit geval uitvoeren?

# Preliminary results on reserve level

Yearly change of basal area for the studied forest reserves

```{r}
calculate_DAS_indicator(data_dendro) %>%
  ggplot(aes(x = forest_reserve, y = d_res)) + geom_bar(stat = "identity")
```

Yearly proportional change of the basal area for each species group.

```{r}
calculate_DAS_indicator(data_dendro) %>%
  ggplot(aes(x = forest_reserve, y = d_group, fill = group)) +
  geom_bar(stat = "identity", position = "dodge")
```
